Monte Carlo tree search compared to A\(^*\) in airspace configuration decision problem

### 2018-05-18 article/article.tex
(+) new figure with caption

The dynamic airspace configuration problem requires a model of the airspace.
In~\cite{sergeeva2017dynamic} or~\cite{treimuth2016branch} the airspace
is modelled via a graph. In those graphs, vertices represent elementary modules
and an edge links two adjacent modules. In~\cite{treimuth2016branch}, to be
able to produce a sequence of configurations, the graph is time dependent.
An other way to model the problem is to use a constrained set of
configurations as in~\cite{gianazza2010forecasting} and~\cite{bloem2010dynamic}.
This way any configuration will match specified requirements, which can be
qualified as hard constraints.

@@

In this paper, each partition scheme is based upon the number of aircraft in
each
elementary module and the cost of transition from the previous partitioning. The
available partitions come from a predetermined set containing feasible
partitioning.

@@

\subsubsection{Time step}
Since the overall goal is to compute a temporal sequence of partitions over a
day, the day has to be sampled. This sampling in represented as the temporal
difference between a node and its children in the trees mentioned in the
previous section.

@@

\subsection{Heuristic}
To use effectively the A\(^*\) algorithm, a heuristic is needed. Using the null
heuristic (\(\forall u \in \mathcal{N}, h(u) = 0\)) revealed to be inefficient
considering the branching factor. To build a minimal heuristic, the limitations
introduced in~\ref{sssec:transitions} are not considered, neither those associated
to the previous partition (split, merge or transfer). This allows to generate
the sequence of the best partitions among all possible at each time step.

@@

Considering paragraph~\ref{ssec:theoretical_performances}, the MCTS should
eventually converge to the optimal value. This result reveals a flaw in the
implementation.
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
      [line width=0.05, mark size=0.1, xlabel=time(s), ylabel=cost]
      \addplot+ table {data/longterm.data};
      \addplot+[dashed, domain=0:10] {1268};
    \end{axis}
  \end{tikzpicture}
  \caption{The continuous line shows the total cost of a path computed by the
    MCTS algorithm. The x-axis shows the time allowed to the algorithm. The
    dashed curve is 15\% of the optimal cost found with the \(A^*\)
  algorithm.}\label{fig:longterm}
\end{figure}
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
      [line width=0.05, mark size=0.1, xlabel=time(s), ylabel=cost]
      \addplot+ table {data/shortterm.data};
      \addplot+[dashed, domain=0:2] {1268};
    \end{axis}
  \end{tikzpicture}
  \caption{}\label{fig:shortterm}
\end{figure}

>>>>

The dynamic airspace configuration problem requires a model of the airspace.
In~\cite{sergeeva2017dynamic} or~\cite{treimuth2016branch} the airspace
is modelled via a graph. In those graphs, vertices represent elementary modules
and an edge links two adjacent modules. In~\cite{treimuth2016branch}, to be
able to produce a sequence of configurations, the graph is time dependent.
An other way to model the problem is to use a constrained set of
configurations as in~\cite{bloem2010dynamic}.
This way any configuration will match specified requirements, which can be
qualified as hard constraints.

@@

In this paper, each partition scheme is based upon the number of aircraft in
each
elementary module and the cost of transition from the previous partitioning.
The set of available parititions can be computed from the set of elementary
modules and a context which contains all available ATC sectors.

@@

\subsubsection{Time step}
Since the overall goal is to compute a temporal sequence of partitions over a
day, the day has to be sampled. In the model tree, a timestep separates a node
from its children. Practically, a timestep is one minute.

@@

\subsection{Heuristic}
To use effectively the A\(^*\) algorithm, a heuristic is needed. Using the null
heuristic (\(\forall u \in \mathcal{N}, h(u) = 0\)) -- and thus using Dijkstra
algorithm -- revealed to be inefficient
considering the branching factor. Let \(k \in \mathbb{N}\), \(u\) a node
selected at time \(k\).
The heuristic \(h(u)\) gives the cost of the path starting from \(u\) to a
leaf which is reached at a time \(t_f\).
The heuristic is built without
considering transitions (transition cost and~\ref{sssec:transitions}).
Let \(P^*_i\) be the partition with minimal cost at time \(i\), the heuristic is
then
\begin{equation}
  h(u) = \sum_{i=k+1}^f C(P^*_i, t_i)
\end{equation}
This allows to generate
the sequence of the best partitions among all possible at each time step.

@@

Considering paragraph~\ref{ssec:theoretical_performances}, the MCTS should
eventually converge to the optimal value. This result reveals a flaw in the
implementation.
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
      [line width=0.05, mark size=0.1, xlabel=time(s), ylabel=cost]
      \addplot+ table {data/longterm.data};
      \addplot+[dashed, domain=0:10] {1268};
    \end{axis}
  \end{tikzpicture}
  \caption{The continuous line shows the total cost of a path computed by the
    MCTS algorithm. The x-axis shows the time allowed to the algorithm. The
    dashed curve is 15\% of the optimal cost found with the \(A^*\)
  algorithm.}\label{fig:longterm}
\end{figure}
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
      [line width=0.05, mark size=0.1, xlabel=time(s), ylabel=cost]
      \addplot+ table {data/shortterm.data};
    \end{axis}
  \end{tikzpicture}
  \caption{Monte Carlo tree search called with varying time allowed to compute
  each step, with a depth of 40 and \(\theta = 11\). The cost seems to be
  approximately stable from \(t=0.5\)s.}\label{fig:shortterm}
\end{figure}
